---
editor: 
  markdown: 
    wrap: 72
execute:
  echo: false
  results: hide
  message: false
  warning: false
---

# Data

## Technical description

The dataset contains tracking outputs from ten videos of mice placed
individually in the same circular arena. Each frame includes x/y
coordinates and likelihood values for several body parts: nose, ears,
head area, tail base, and tail tip. Videos are around 45,000–51,000
frames each (roughly 25–28 minutes at 30 fps).

All tracking files were collected directly from our lab’s DeepLabCut
pipeline.

Overall, the data are consistent in format and length across videos,
which makes them easy to compare. The only issue was that likelihood for
the tail tip was sometimes lower than other body parts, which is common
in freely moving animals. Aside from that, the data quality was high.

```{r}

library(tidyverse)

files <- list.files("data", pattern = "\\.csv$", full.names = TRUE)

load_pose <- function(path) {
  
  raw <- read_csv(path, col_names = FALSE, show_col_types = FALSE)
  
  # DeepLabCut stores headers in rows 1–3:
  # Row 1 = scorer, Row 2 = bodypart, Row 3 = coordinate
  header <- raw[1:3, ]
  data <- raw[-c(1:3), ]
  
  bodyparts <- as.character(header[2, ])
  coords <- as.character(header[3, ])
  new_names <- paste0(bodyparts, "_", coords)
  
  colnames(data) <- new_names
  
  data <- data |> mutate(across(everything(), as.numeric))
  
  data$video_id <- basename(path)
  data$frame <- seq_len(nrow(data))
  
  data
}

pose_all <- map_df(files, load_pose)

glimpse(pose_all)
```

## Missing Value Analysis

```{r}
library(tidyverse)

# Raw DLC variables ending with _x, _y, _likelihood
raw_vars <- names(pose_all) |>
  str_subset("_(x|y|likelihood)$")

missing_by_vid_var <- pose_all |>
  group_by(video_id) |>
  summarize(
    across(all_of(raw_vars), ~ mean(is.na(.))),
    .groups = "drop"
  ) |>
  pivot_longer(
    cols = -video_id,
    names_to = "variable",
    values_to = "prop_missing"
  )

ggplot(missing_by_vid_var,
       aes(x = video_id, y = variable, fill = prop_missing)) +
  geom_tile() +
  geom_text(
    aes(label = ifelse(prop_missing > 0,
                       scales::percent(prop_missing, accuracy = 0.01),
                       "")),
    size = 3
  ) +
  scale_fill_viridis_c(labels = scales::percent,
                       name = "Percent\nmissing") +
  labs(
    title = "Missingness by video and variable",
    x = "Video file",
    y = "Variable"
  ) +
  theme_minimal(base_size = 14) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

I checked for missing values across all body parts and videos. According
to the heatmap, there were no missing values at all.

```{r}
library(tidyverse)

raw_vars <- names(pose_all) |>
  str_subset("_(x|y|likelihood)$")

zero_by_vid_var <- pose_all |>
  group_by(video_id) |>
  summarize(
    across(all_of(raw_vars), ~ mean(. == 0, na.rm = TRUE)),
    .groups = "drop"
  )|>
  pivot_longer(
    cols = -video_id,
    names_to = "variable",
    values_to = "prop_zero"
  )

ggplot(zero_by_vid_var,
       aes(x = video_id, y = variable, fill = prop_zero)) +
  geom_tile() +
  geom_text(
    aes(label = ifelse(prop_zero > 0,
                       scales::percent(prop_zero, accuracy = 0.1),
                       "")),
    size = 3
  ) +
  scale_fill_viridis_c(
    labels = scales::percent,
    name   = "Percent\nzeros"
  ) +
  labs(
    title = "Fraction of zero values by video and variable",
    x = "Video file",
    y = "Variable"
  ) +
  theme_minimal(base_size = 14) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

A second plot showed the fraction of zeros, which was extremely small
(0.1–0.5%).

```{r}
library(tidyverse)

likelihood_vars <- names(pose_all) |>
  str_subset("likelihood$")

# Compute fraction of frames with likelihood > 0.9
high_conf_by_vid <- pose_all |>
  group_by(video_id) |>
  summarize(
    across(all_of(likelihood_vars), 
           ~ mean(. > 0.9, na.rm = TRUE)),
    .groups = "drop"
  ) |>
  pivot_longer(
    cols = -video_id,
    names_to = "variable",
    values_to = "prop_high"
  )

ggplot(high_conf_by_vid,
       aes(x = video_id, y = variable, fill = prop_high)) +
  geom_tile() +
  geom_text(
    aes(label = scales::percent(prop_high, accuracy = 0.1)),
    size = 3
  ) +
  scale_fill_viridis_c(
    labels = scales::percent,
    name = "Likelihood > 0.9"
  ) +
  labs(
    title = "Fraction of frames with likelihood > 0.9",
    x = "Video file",
    y = "Body part"
  ) +
  theme_minimal(base_size = 14) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

Next, I examined detection confidence. Most likelihood values were above
0.9 for nearly all body parts and videos, except the tail tip. Because
the majority of points were reliable, and filtering by likelihood would
remove too much data, I kept all frames for analysis.

Together, these checks confirmed that the dataset is complete and
stable, with no missingness concerns.
