---
editor: 
  markdown: 
    wrap: 72
---

# Data

## Technical description

The dataset consists of ten DeepLabCut pose-tracking CSV files. Each
includes multiple tracked body parts (nose, ears, tail base, etc.) with
three measurements per part: \*\*x\*\*, \*\*y\*\*, and
\*\*likelihood\*\*. The first three rows of each CSV contain DeepLabCut
header information (scorer, bodypart name, coordinate type). These rows
are removed, and columns are rebuilt into the format

```{r}

library(tidyverse)

# Load all CSV files
files <- list.files("data", pattern = "\\.csv$", full.names = TRUE)

load_pose <- function(path) {
  
  # Read entire CSV with NO column names
  raw <- read_csv(path, col_names = FALSE, show_col_types = FALSE)
  
  # DeepLabCut stores headers in rows 1–3:
  # Row 1 = scorer, Row 2 = bodypart, Row 3 = coordinate
  header <- raw[1:3, ]
  data <- raw[-c(1:3), ]
  
  # Build proper column names
  bodyparts <- as.character(header[2, ])
  coords <- as.character(header[3, ])
  new_names <- paste0(bodyparts, "_", coords)
  
  colnames(data) <- new_names
  
  # Convert everything to numeric
  data <- data %>% mutate(across(everything(), as.numeric))
  
  # Add metadata
  data$video_id <- basename(path)
  data$frame <- seq_len(nrow(data))
  
  data
}

pose_all <- map_df(files, load_pose)

glimpse(pose_all)
```

```{r}
library(tidyverse)

# Raw DLC variables: end with _x, _y, _likelihood
raw_vars <- names(pose_all) %>%
  str_subset("_(x|y|likelihood)$")

missing_by_vid_var <- pose_all %>%
  group_by(video_id) %>%
  summarize(
    across(all_of(raw_vars), ~ mean(is.na(.))),
    .groups = "drop"
  ) %>%
  pivot_longer(
    cols = -video_id,
    names_to = "variable",
    values_to = "prop_missing"
  )

ggplot(missing_by_vid_var,
       aes(x = video_id, y = variable, fill = prop_missing)) +
  geom_tile() +
  geom_text(
    aes(label = ifelse(prop_missing > 0,
                       scales::percent(prop_missing, accuracy = 0.01),
                       "")),
    size = 3
  ) +
  scale_fill_viridis_c(labels = scales::percent,
                       name = "Percent\nmissing") +
  labs(
    title = "Missingness by video and variable",
    x = "Video file",
    y = "Variable"
  ) +
  theme_minimal(base_size = 14) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r}
library(tidyverse)

# Use only the raw DeepLabCut variables (skip frame, video_id, dx, speed, cluster, etc.)
raw_vars <- names(pose_all) %>%
  str_subset("_(x|y|likelihood)$")

zero_by_vid_var <- pose_all %>%
  group_by(video_id) %>%
  summarize(
    across(all_of(raw_vars), ~ mean(. == 0, na.rm = TRUE)),
    .groups = "drop"
  ) %>%
  pivot_longer(
    cols = -video_id,
    names_to = "variable",
    values_to = "prop_zero"
  )

ggplot(zero_by_vid_var,
       aes(x = video_id, y = variable, fill = prop_zero)) +
  geom_tile() +
  geom_text(
    aes(label = ifelse(prop_zero > 0,
                       scales::percent(prop_zero, accuracy = 0.1),
                       "")),
    size = 3
  ) +
  scale_fill_viridis_c(
    labels = scales::percent,
    name   = "Percent\nzeros"
  ) +
  labs(
    title = "Fraction of zero values by video and variable",
    x = "Video file",
    y = "Variable"
  ) +
  theme_minimal(base_size = 14) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r}
library(tidyverse)

# Select only likelihood columns
likelihood_vars <- names(pose_all) %>% 
  str_subset("likelihood$")

# Compute fraction of frames with likelihood > 0.9
high_conf_by_vid <- pose_all %>%
  group_by(video_id) %>%
  summarize(
    across(all_of(likelihood_vars), 
           ~ mean(. > 0.9, na.rm = TRUE)),
    .groups = "drop"
  ) %>%
  pivot_longer(
    cols = -video_id,
    names_to = "variable",
    values_to = "prop_high"
  )

# Plot heatmap
ggplot(high_conf_by_vid,
       aes(x = video_id, y = variable, fill = prop_high)) +
  geom_tile() +
  geom_text(
    aes(label = scales::percent(prop_high, accuracy = 0.1)),
    size = 3
  ) +
  scale_fill_viridis_c(
    labels = scales::percent,
    name = "Likelihood > 0.9"
  ) +
  labs(
    title = "Fraction of frames with likelihood > 0.9\nfor each body part and video",
    x = "Video file",
    y = "Body part"
  ) +
  theme_minimal(base_size = 14) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```
